\section{Introduction}
\label{sec:introduction}

Compute-in-memory (CIM) and processing-in-memory (PIM) architectures are widely motivated by the energy cost of moving data between memory and arithmetic units in edge inference systems \citep{horowitz2014computing,peng2020dnnneurosim}. In the present work, however, the relevant question is not generic AI acceleration. It is whether emerging organic optoelectronic synaptic devices can support useful image-classification accuracy under realistic non-idealities while still preserving the analog vector-matrix multiplication advantages that make CIM attractive in the first place.

Organic optoelectronic and electrochemical synaptic devices are compelling because they offer multilevel conductance tuning, optical sensitivity, and ultra-low static power \citep{xu2025emerging,guo2024organic}. Yet most device-facing studies stop at idealized network validation. Variability, converter precision, retention drift, and deployment-time calibration are often omitted or only discussed qualitatively \citep{zeng2023organicmemristor,jung2024organicfilaments}. As a result, there is still a gap between the quantities that materials researchers can measure early in a device program and the quantities that system researchers need in order to judge whether a given hardware stack is likely to fail at the model level.

That gap is especially visible for organic-device simulation. The best-known variability-aware organic precursor is the Monte Carlo study of Alibart \emph{et al.}, which evaluated a simple organic-memristive perceptron under repeated trials \citep{alibart2016physical}. General CIM simulators such as DNN+NeuroSim and MemTorch have subsequently matured for inorganic resistive memories \citep{peng2020dnnneurosim,memtorch}, but they do not directly answer how organic optoelectronic assumptions propagate through hybrid analog/digital deployment choices in modern convolutional and transformer backbones.

This paper addresses that gap through a full-stack, hardware-aware simulation framework centered on organic optoelectronic synaptic transistors for edge-vision inference. The framework starts from crossbar-aware mapping rules, injects quantization, cycle-to-cycle (C2C) and device-to-device (D2D) variability, ADC distortion, retention drift, and physical-front-end effects directly into the forward path, and supports both literature-anchored priors and later measured-device substitution through a shared profile interface. The framework is evaluated across ResNet-18, ConvNeXt-Tiny, and Tiny-ViT-5M so that both convolutional and hybrid transformer deployment regimes are represented.

The main contributions are fourfold. First, we provide a PyTorch-based materials-to-system simulation workflow for organic optoelectronic CIM under joint quantization, variability, retention, and peripheral constraints. Second, we extend organic-device robustness analysis beyond small perceptrons to contemporary vision backbones, including a hybrid transformer deployment. Third, we integrate multiple non-idealities in one stack, including nonlinear write surrogates and state-dependent noise stress tests. Fourth, we structure the framework so that measured laboratory profiles can later replace literature priors without requiring code-path changes.

The rest of the paper is organized as follows. Section~2 reviews related work on organic neuromorphic devices, CIM simulation frameworks, and hardware-aware training. Section~3 presents the hybrid mapping strategy, analog crossbar model, variability and retention formulations, and the energy-estimation methodology. Section~4 summarizes the experimental setup, including the model suite, datasets, and experiment matrix. Section~5 reports the cross-dataset results, retention behavior, front-end compensation results, and physically extended stress tests. Section~6 discusses the dominant bottlenecks, architectural differences, and materials-facing implications. Section~7 concludes with the scope and limits of the present first-order behavioral framework.
