% !TeX root = ../main.tex
\section{Methodology}
\label{sec:methodology}

\subsection{Hybrid Analog/Digital Deployment}
\label{subsec:hybrid-deployment}

We adopt a hybrid analog/digital inference stack: dense linear operators execute on differential-pair crossbars; control-heavy operators remain digital. Concretely, the analog partition contains patch embeddings, query/key/value (Q/K/V) and output projections, and feed-forward layers; the digital partition contains softmax, layer normalization, GELU activation, positional encoding, residual additions, and the class token. The system-level split and the operator-level rationale are summarized in the Supplementary Information following established practice in heterogeneous CIM accelerators \citep{wu2023bwq,wang2024epim,ge2024allspark,kim2025hemlet,lin2024hardsea}.

Under this mapping, approximately 88\% of parameters are assigned to analog execution, whereas roughly 60\% of the estimated energy remains in the digital domain under the present placeholder constants, largely because of attention operations. This split reflects a basic architectural constraint: analog crossbars are well suited to weight-stationary matrix multiplication, but much less suited to dynamic, input-dependent computations \citep{kim2025hemlet}. Recent heterogeneous CIM accelerators for Vision Transformers follow the same logic, keeping attention digital even as linear projections move to analog arrays \citep{lin2024hardsea,wang2024epim,ge2024allspark}. The present organic device literature likewise demonstrates regular static array programming more naturally than fully dynamic interaction engines \citep{gebregiorgis2023organiccim,zhang2025opect}. Layer-wise mapping ablations that vary the analog--digital boundary are supported by the framework's modular operator assignment, but they are deferred here so that the present study can isolate the impact of device non-idealities on a single, widely adopted hybrid partition.

\subsection{Modeling Physical Non-Idealities}
\label{subsec:modeling-nonidealities}

The framework injects quantization, cycle-to-cycle (C2C) and device-to-device (D2D) variability, ADC distortion, retention drift, and nonlinear-write surrogates directly into the forward path. Each analog-mapped weight is decomposed into positive and negative branches, mapped to the conductance window $[G_{\min}, G_{\max}]$, and quantized to $n_{\mathrm{states}}$ levels via a straight-through estimator (STE) quantizer $Q_{n}(\cdot)$. Differential readout recovers $G_{\mathrm{eff}}=\tilde{G}^{+}-\tilde{G}^{-}$, and the downstream digital weight is
\begin{equation}
    \hat{W}_{ij}=s_{\ell}\,G_{\mathrm{eff},ij},
    \qquad
    s_{\ell}=
    \begin{cases}
        \dfrac{w_{\max}}{G_{\max}-G_{\min}}, & \text{standard calibration} \\[6pt]
        \dfrac{w_{\max}}{\max_{i,j}|G_{\mathrm{eff},ij}|+\epsilon}, & \text{retention-time recalibration}
    \end{cases}
\label{eq:scale-recovery}
\end{equation}
with layer index $\ell$ omitted. Unless otherwise stated, we use the standard-calibration branch; the retention-recalibration branch is used only for the V8 retention-drift experiments (Section~\ref{subsec:retention-drift}). The recovered weight step is $\Delta_{\hat{W}}=s_{\ell}(G_{\max}-G_{\min})/(n_{\mathrm{states}}-1)$, so perturbations smaller than half a step are masked. The differential-pair scheme suppresses common-mode noise \citep{wei2020voltagedifferential,kim2024sttmram}.

Hardware-aware training (HAT) keeps forward perturbations active during optimization while back-propagating through a straight-through estimator. Standard HAT fixes one D2D mismatch field $M_{0}$,
\begin{equation}
    \theta_{\mathrm{std}}^{\star}=\arg\min_{\theta}\frac{1}{N}\sum_{n=1}^{N}\ell\!\left(f(x_{n};\theta,M_{0}),y_{n}\right),
\label{eq:hat-standard}
\end{equation}
\paragraph{Ensemble HAT as distribution matching.}
Whereas Standard HAT fixes one D2D mismatch realization $M_0$ (Eq.~\ref{eq:hat-standard}), Ensemble HAT treats the mismatch map as a random variable resampled each epoch:
\begin{equation}
    \theta_{\text{ens}}^{\star} = \arg\min_{\theta}\;\mathbb{E}_{M\sim\mathcal{D}_{\text{D2D}}}\Bigl[\frac{1}{N}\sum_{n=1}^{N}\ell\bigl(f(x_n;\theta,M),y_n\bigr)\Bigr],
    \label{eq:hat-ensemble-distribution}
\end{equation}
where the expectation is Monte-Carlo estimated with one fresh draw $M^{(e)}\sim\mathcal{D}_{\text{D2D}}$ held constant across all mini-batches in epoch $e$. A second-order Taylor expansion of Eq.~\ref{eq:hat-ensemble-distribution} in the mismatch amplitude (Supplementary Note S-Theory) reveals that Ensemble HAT introduces an \emph{implicit regularizer} proportional to $\sigma_{\text{D2D}}^2\sum_i \theta_i^2(\partial\ell/\partial\theta_i)^2$ --- a Fisher-weighted gradient-squared penalty that pushes the optimizer toward flat regions of the loss landscape in the $M$-direction. This mechanistic interpretation explains the empirical fresh-instance transferability without requiring additional hyperparameters: the regularization strength is set by the physical D2D variance rather than by a tunable dropout rate. The epoch-level resampling discipline (as opposed to per-batch) is load-bearing: it gives the optimizer enough consistent signal to fit each instance before resampling, enforcing a ``learn-to-adapt'' regime analogous to domain randomization \citep{tobin2017domain}.
Fresh-instance transfer evaluates on unseen draws $M'\sim p(M)$:
\begin{equation}
    \mathrm{Acc}_{\mathrm{fresh}}=\mathbb{E}_{M'\sim p(M)}\!\left[\mathrm{Acc}\!\left(f(\cdot;\theta_{\mathrm{ens}}^{\star},M')\right)\right].
\label{eq:fresh-instance}
\end{equation}
Each fresh array corresponds to an i.i.d. Gaussian draw $M' \sim \mathcal{N}(0, \sigma_{\mathrm{D2D}}^2)$ held constant over the evaluation pass, distinct from the cycle-to-cycle perturbation that re-samples per forward pass.
For the canonical Ensemble HAT transfer result, each fresh-instance mean is itself the mean of five forward-pass Monte Carlo evaluations, and the reported $\pm 1.54$\,pp spread is the standard deviation across the ten fresh-instance means rather than across all fifty forward passes pooled together.

\subsection{Training Protocol}
\label{subsec:training-protocol}

All Tiny-ViT experiments are trained with AdamW ($lr = 5\times10^{-4}$, weight decay $0.05$) for 100 epochs using a batch size of 64 and a cosine annealing learning-rate schedule ($T_{\max}=100$). Images are resized to $224\times224$, augmented with random horizontal flips, and normalized by dataset statistics. The default analog programming resolution is $n_{\mathrm{states}}=16$ conductance levels. Experiment V1 uses standard digital training; V2 uses hybrid layers without noise; V3 uses hybrid layers with fixed D2D mismatch ($\sigma_{\mathrm{D2D}}=10\%$) active during training but no per-forward C2C resampling; V4--V7 employ full hardware-aware training in which fixed D2D mismatch is resampled at the start of every epoch and C2C noise ($\sigma_{\mathrm{C2C}}=5\%$ or $10\%$) is drawn on each forward pass. V6 additionally enables the physical photoresponse front end, and V7 evaluates post-training retention decay. Complete experiment settings are documented in Supplementary Table~S1.

First-order nonlinear-write and double-exponential retention surrogates are supported, with parameters anchored to measured dinaphtho-thieno-thiophene (DNTT) transients \citep{vincze2025dualplasticity}. Nonlinear write modifies the STE backward pass by scaling the straight-through gradient with a state-dependent factor proportional to the present conductance magnitude raised to $NL-1$ (Supplementary Eq.~S2 gives the exact piecewise form)\footnote{The absence of an explicit $NL$ prefactor is intentional: this is a behavioral proxy for position-dependent update difficulty, not the strict derivative of a power-law $f(G)=G^{NL}$.}; ideal write is recovered when $NL_{\mathrm{LTP}}=1$ (long-term potentiation) and $|NL_{\mathrm{LTD}}|=1$ (long-term depression). For optoelectronic sensing, we incorporate a physical front-end model (V6) that first applies inverse-gamma preprocessing to the normalized input $X \in [0,1]$,
\begin{equation}
    P_{\mathrm{in}} = X^{1/\gamma_{\mathrm{phys}}},
\label{eq:inverse-gamma}
\end{equation}
and then propagates it through a photocurrent response model,
\begin{equation}
    I_{\mathrm{photo}}
    =
    \alpha P_{\mathrm{in}}^{\gamma_{\mathrm{phys}}}
    +
    I_{\mathrm{dark}}
    +
    \varepsilon_{\mathrm{shot}}
    \approx
    \alpha X + I_{\mathrm{dark}} + \varepsilon_{\mathrm{shot}},
\label{eq:frontend-photoresponse}
\end{equation}
with $\mathrm{Var}[\varepsilon_{\mathrm{shot}}] \propto I_{\mathrm{photo}}$. The proportionality coefficient is absorbed into the per-profile front-end calibration, so its exact value (governed by photodetector quantum efficiency and integration time) is implicit in the reported results. The uncompensated form ($P_{\mathrm{in}}=X$) and the front-end theory derivation are given in the Supplementary Information. The simulated photocurrent is finally renormalized on a per-sample basis,
\begin{equation}
    X_{\mathrm{norm}}
    =
    \frac{I_{\mathrm{photo}}-I_{\min}}{I_{\max}-I_{\min}+\epsilon},
\label{eq:frontend-renorm}
\end{equation}
before dataset normalization. The energy model uses representative edge-node parameters.


The full weight-to-conductance mapping pipeline is illustrated in the Supplementary Information.

\subsection{Profile-Driven Substitution Interface}
\label{subsec:profile-interface}

The profile interface packages technology-specific assumptions into a replaceable JSON parameter bundle $(\sigma_{\mathrm{D2D}}, \sigma_{\mathrm{C2C}}, n_{\mathrm{states}}, G_{\min}, G_{\max}, NL_{\mathrm{LTP}}, NL_{\mathrm{LTD}}, \tau_{1}, \tau_{2}, A_{0}, \gamma_{\mathrm{phys}}, I_{\mathrm{dark}}, \ldots)$. The deployment evaluation holds the digital backbone and training recipe fixed while substituting the profile into the same mixed-signal inference path, enabling literature priors today and measured-device characterizations later without code changes. Representative profile fields and provenance are documented in the Supplementary Information.

\subsection{Sensitivity and Energy Metrics}
\label{subsec:sensitivity-energy}

For the joint D2D--ADC sweep, we quantify first-order factor importance with Sobol indices 
\begin{equation}
    S_{i} = \frac{\mathrm{Var}_{X_{i}}\!\left(\mathbb{E}[Y \mid X_{i}]\right)}{\mathrm{Var}(Y)},
\label{eq:sobol-first-order}
\end{equation}
estimated directly from the $7 \times 9$ grid of Monte Carlo means as $\mathrm{Var}(\mathrm{mean}_{\text{grid}}[X_{i}]) / \mathrm{Var}(Y)$, with MC noise realizations drawn independently per grid point and per architecture (seeds are not shared across the grid or across models). Indices are reported over both the full grid and the operational subregion. Energy is estimated via a first-order analytical model described in the Supplementary Methods from operator counts and literature-proxy per-operation constants ($E_{\text{cell}}=100$~fJ, $E_{\text{ADC},8b}=25$~fJ, $E_{\text{DAC},8b}=30$~fJ); these are analytical placeholders, not measured silicon values, and any reported energy ratios are contingent on these assumptions. The numbers are intended for relative comparison rather than routed-circuit prediction.

