% =============================================================================
% Acknowledgements, Dedication, and Declaration
% =============================================================================

\chapter*{Dedication}
\addcontentsline{toc}{chapter}{Dedication}

\vspace*{2cm}
\begin{center}
\textit{To my family,\\
whose unwavering belief in me has been the quiet foundation upon which\\
every ambition of mine has been built.}
\end{center}

\vspace*{1cm}
\begin{center}
\textit{This work is dedicated to all those who reminded me,\\
in moments of doubt, that the pursuit of knowledge is never walked alone.}
\end{center}

\clearpage

% =============================================================================
\chapter*{Acknowledgements}
\addcontentsline{toc}{chapter}{Acknowledgements}

First and foremost, I owe an immense debt of gratitude to my advisor, [Advisor Name]  
% TODO: User requested to keep placeholders for now, whose intellectual generosity and patience have shaped not only this thesis but my entire approach to research. [Advisor Name] granted me the freedom to explore unconventional directions while providing the steady guidance necessary to bring those explorations to fruition. I am equally thankful to the members of my dissertation committee---[Committee Member 1], [Committee Member 2], and [Committee Member 3]---for their insightful critiques, challenging questions, and the breadth of perspective they brought to this work.

This research would not have been possible without the remarkable collaborators who contributed their expertise at critical junctures. I am deeply grateful to [Collaborator Name] for invaluable discussions on device physics and for helping frame the physical constraints that grounded our architectural decisions. [Collaborator Name] provided rigorous code review and software engineering insight that transformed a collection of experiments into a reproducible research pipeline. I also thank [Collaborator Name] for their contributions to cross-layer simulation methodology and for countless whiteboard sessions that clarified my thinking.

The computational experiments presented in this thesis were enabled by generous access to high-performance computing resources. I gratefully acknowledge the NVIDIA Apamayo cluster and the [Institution] Research Computing Center for providing the GPU and server infrastructure that made large-scale neural architecture search and co-design studies feasible. I also extend my thanks to the system administrators and support staff who maintained these resources with professionalism and responsiveness.

Financial support for this work was provided by [Funding Agency 1] under Grant No.~[Grant Number], [Funding Agency 2] under Grant No.~[Grant Number], and [Fellowship Name]. Their investment in fundamental research made this dissertation possible, and I am honored to have been supported by their vision.

Finally, to my family, friends, and partner: thank you for bearing with the absences, the late nights, and the occasional obsession. Your emotional support was not incidental to this work---it was essential. You reminded me that there is a world beyond the terminal, and that perspective kept me whole.

\clearpage

% =============================================================================
\chapter*{Declaration}
\addcontentsline{toc}{chapter}{Declaration}

I hereby declare that this thesis, titled ``[Thesis Title],'' is my own original work and has not been submitted for any other degree or qualification at this or any other institution. All sources used in the preparation of this thesis have been duly acknowledged. The research reported herein was conducted under the supervision of [Advisor Name] at [Institution Name]. Where the work of others is referenced or built upon, proper attribution has been given. Any errors or omissions remain my own.

\vspace{2cm}
\noindent
[Author Name]\\
[City, Country]\\
[Month Year]

% =============================================================================
\endinput
