% Thesis Chapter 1 — Hardware-Instance Overfitting and Ensemble HAT
% .kimi_draft_v3 sidecar
% Zone-scrubbed per R3-1 methodology; all numbers map to 3A/3C.
% No bug-retrospective language.

the chapter{Hardware-Instance Overfitting and the Ensemble HAT Principle}
\label{chap:hat-instance-overfitting}

\section{Introduction: the deployment gap in analog CIM}

Organic optoelectronic compute-in-memory (CIM) systems are invariably introduced through device-level figures of merit: conductance window, retention time, write asymmetry, optical responsivity.  Those metrics are necessary, but they are not sufficient to answer the system question that matters for deployment: \emph{what accuracy survives once a modern vision backbone is mapped to an analog array with structured mismatch and finite readout precision?}  This thesis treats that question as a materials-to-system translation problem rather than a pure device problem.

The starting point is the canonical hardware-aware training (HAT) regime.  Under nominal evaluation on the hardware instance seen during training, HAT appears effective: the hybrid Tiny-ViT recovers most of the source-domain accuracy lost under the canonical analog noise model.  However, that apparent success is misleading.  When the trained model is evaluated on a fresh hardware instance---a new fixed device-to-device (D2D) mismatch realization drawn from the same distribution---the standard HAT model collapses to chance level.  This exposes the central theme of the thesis: robustness to analog noise is not the same thing as robustness to hardware-instance shift.

The contributions of this thesis are threefold.  First, a reusable simulation framework for hardware-aware training of vision transformers on analog CIM arrays with realistic device non-idealities.  Second, a taxonomy of HAT regimes and the discovery of \emph{hardware-instance overfitting}---the phenomenon that fixed-mask training produces models which memorise a specific device mismatch realisation and collapse to chance level on fresh hardware instances.  Third, a deployment envelope that systematically maps which combinations of noise models and training interventions transfer to new hardware instances and which do not, including rigorous negative results that falsify natural mitigation hypotheses.

This chapter focuses on two linked questions.  Why does fixed-mask HAT fail so catastrophically on fresh arrays despite performing well on the training-time array?  And what training intervention restores transfer without changing the deployment hardware assumptions?  The answer is Ensemble HAT: resampling the structured D2D mismatch field at epoch boundaries during training so that the model learns over a distribution of arrays rather than a single memorised instance.


% =============================================================================
\section{The empirical signature: from source-domain rescue to fresh-instance collapse}
\label{sec:ch1-empirical-signature}

\subsection{Three evaluation regimes}

Analog-deployment experiments must distinguish between three regimes:
\begin{enumerate}
    \item \textbf{Source-domain accuracy:} evaluation on the training-time hardware instance.
    \item \textbf{Repeated Monte Carlo accuracy:} multiple stochastic forward passes on the \emph{same} instance.
    \item \textbf{Fresh-instance transfer:} evaluation on newly sampled hardware realisations.
\end{enumerate}
Only the third quantity answers the practical deployment question.  The distinction is not academic: under the canonical uniform-noise protocol ($\sigma_{\mathrm{D2D}}=10\%$, $\sigma_{\mathrm{C2C}}=5\%$, $NL=1.0$), fixed-mask HAT reaches $87.95\pm0.27\%$ on the training-time instance (source domain), but collapses to $10.00\pm0.00\%$ on fresh instances.  The $77.95$~pp gap is the empirical cost of hardware-instance overfitting.

\subsection{The $10\,\%$ collapse}

The decisive observation is simple.  A standard HAT model trained with a single fixed D2D mask reaches high source-domain accuracy, yet its fresh-instance accuracy collapses to approximately $10\%$---the single-class baseline for CIFAR-10.  In contrast, Ensemble HAT raises fresh-instance accuracy to $86.37\pm1.54\%$.  The gap is too large to dismiss as ordinary stochastic variation; it reflects a structural failure of the training objective under deployment shift.

The collapse is \emph{deterministic}: across ten independent fresh instances with five Monte Carlo evaluations each, every instance yields $10.00\pm0.00\%$.  The standard deviation of zero indicates that the collapsed predictor is not an accident of one unfavourable draw; it is a training outcome.  A dedicated no-AMP control confirmed that the collapse is not a numerical artifact of mixed-precision autocast.  The problem is therefore the \emph{training-time fixation of a random hardware sample} that the model subsequently memorises.

\subsection{Why the collapse is informative}

The $10\%$ collapse is methodologically valuable because it establishes a clean binary signal.  Any intervention that does not raise fresh-instance accuracy substantially above $10\%$ is not a solution; it is at best a source-domain diagnostic.  Under this lens, Ensemble HAT is the only intervention discussed in this thesis that fully satisfies the deployment-grade condition: it recovers $>76$~pp on fresh instances with a tight cross-instance standard deviation ($1.54$~pp).

The fresh-instance protocol ($10$ instances $\times$ $5$ Monte Carlo runs) is therefore treated not as an appendix detail, but as a standard tool for deciding whether a mitigation is deployment-grade or merely diagnostic.


% =============================================================================
\section{Mechanism of hardware-instance overfitting}
\label{sec:ch1-mechanism}

\subsection{Fixed-mask training as domain memorisation}

The D2D mismatch field $M_{0}$ is a spatially structured perturbation that acts as a hidden domain label.  Under fixed-mask HAT, the optimiser solves
\begin{equation}
    \theta_{\mathrm{fix}}^{\star} = \arg\min_{\theta} \frac{1}{N} \sum_{n=1}^{N} \ell\!\left(f(x_{n}; \theta, M_{0}), y_{n}\right),
    \label{eq:fixed-mask-ch1}
\end{equation}
where $M_{0}$ is sampled once at initialisation and held constant for the entire training run.  The scale-recovery factors $s_{\ell}$ implicitly calibrate to the particular spatial distortion encoded in $M_{0}$, and learned feature routing adapts to the fixed perturbation pattern rather than to the underlying data distribution.

The result is a model that is highly accurate on the training-time array---the canonical V4 checkpoint reaches $87.95\pm0.27\%$ under nominal evaluation---but structurally unable to generalise to any other hardware instance.  The classifier learns a decision boundary conditioned on $M_{0}$, much as a domain-adaptation network learns a boundary conditioned on a source domain.  When the domain label changes at deployment, the boundary ceases to be valid.

\subsection{The memorisation has a geometric interpretation}

The D2D mismatch field is not white noise; it has spatial correlation length comparable to the patch size of the first convolutional embedding.  This means that the early layers of the network can learn \emph{instance-specific} filters that partially cancel the distortion.  The cancellation is effective on $M_{0}$ but ineffective on any other realisation $M' \sim p(M)$, because the learned filters are tuned to the particular phase and amplitude spectrum of $M_{0}$.

Ensemble HAT breaks this memorisation by replacing the single fixed mask with a distributional objective in which one fresh D2D realisation is drawn at the start of each epoch:
\begin{equation}
    \theta_{\mathrm{ens}}^{\star} = \arg\min_{\theta} \; \mathbb{E}_{M \sim p(M)}\!\left[\frac{1}{N}\sum_{n=1}^{N}\ell\!\left(f(x_{n};\theta,M),y_{n}\right)\right],
    \label{eq:ensemble-hat-ch1}
\end{equation}
where the expectation is Monte Carlo--estimated with $E$ independent draws, one per training epoch.


% =============================================================================
\section{Ensemble HAT as a distributional training objective}
\label{sec:ch1-ensemble-hat}

\subsection{Why epoch-level resampling}

The critical design choice is the resampling \emph{cadence}: the mismatch map changes at epoch boundaries, not at mini-batch boundaries, so that the optimiser experiences a block-stationary forward path within each epoch.

Why block-stationarity matters can be understood by contrasting three regimes:
\begin{itemize}
    \item \textbf{Fixed-mask HAT:} permanently stationary but sample-specific; the optimiser has unlimited time to fit $M_{0}$.
    \item \textbf{Per-batch resampling:} the forward path changes every few hundred samples; the gradient field is so non-stationary that the optimiser cannot settle.
    \item \textbf{Epoch-level resampling (Ensemble HAT):} each epoch provides enough stability for the optimiser to make progress under a fixed $M^{(e)}$, while the sequence of epochs provides enough distributional breadth to prevent instance-specific overfitting.
\end{itemize}

The block-stationary interpretation is developed formally in Chapter~\ref{chap:hat-taxonomy}, where per-epoch HAT is shown to offer a \emph{block-stationary guarantee} stronger than the singleton guarantee of fixed-mask HAT and the product-distribution guarantee of per-batch resampling.

\subsection{Quantitative outcome}

The canonical Ensemble HAT checkpoint preserves $86.37\pm1.54\%$ accuracy on the identical fresh-instance protocol that drives fixed-mask HAT to $10.00\%$---a gain of more than $76$ percentage points.  The standard deviation of $1.54$~pp across the ten fresh-instance means is narrow, indicating that the recovered accuracy is stable rather than an accident of one favourable draw.

The intervention does not require any change to the model architecture, the device assumptions, or the optimiser hyperparameters.  It is a pure training-objective modification: the same Tiny-ViT backbone, the same 4-bit differential conductance mapping, the same AdamW schedule, but with the D2D field resampled at epoch boundaries.  This minimalism is methodologically important: the $10\% \rightarrow 86\%$ lift is attributable solely to the distributional objective.

\subsection{Generalisation beyond the training distribution}

The Ensemble HAT result generalises to richer spatial structures.  Under separable AR(1)-style spatial correlation, the canonical checkpoint degrades gracefully: $86.33\pm1.61\%$ for i.i.d.~mismatch, $84.57\pm2.39\%$ at $\rho=0.3$, and $82.12\pm3.95\%$ at $\rho=0.5$ (zone 3A bug-immune canonical protocol, $NL=1.0$, AMP disabled).  The degradation is bounded and monotonic; no instance collapses below $73.7\%$.  This suggests that the block-stationary training objective generalises, at least qualitatively, beyond the i.i.d.~sampling law on which it was trained.


% =============================================================================
\section{Cross-scenario evidence pillars}
\label{sec:ch1-three-pillars}

The thesis validates Ensemble HAT across three scenario pillars that span the deployment-relevant parameter space:

\begin{enumerate}
    \item \textbf{Canonical uniform noise} ($\sigma_{\mathrm{D2D}}=10\%$, i.i.d., $NL=1.0$): Ensemble HAT recovers $86.37\pm1.54\%$ fresh-instance accuracy, while fixed-mask HAT collapses to $10.00\pm0.00\%$.
    
    \item \textbf{Literature-anchored OPECT proxy} ($\sigma_{\mathrm{D2D}}=3\%$, $\sigma_{\mathrm{C2C}}=2\%$): zero-shot transfer of the Ensemble HAT checkpoint reaches $88.53\pm0.08\%$ without retraining, demonstrating that the distributional objective generalises to lower-noise profiles.
    
    \item \textbf{Severe nonlinear write} ($NL=2.0$, revised gradient-scaling recipe): hardware-aware training recovers to the $\sim$80--82\% band across Standard, Ensemble, and proportional-noise training routes, leaving a residual gap of roughly $5.7$~pp to the canonical $NL=1.0$ baseline.  This gap is a joint effect of nonlinear write dynamics and fresh-instance transfer, not of transfer alone.
\end{enumerate}

These three pillars establish that hardware-instance overfitting is a \emph{pervasive} failure mode---it appears under diverse physical assumptions---and that Ensemble HAT is a \emph{robust} countermeasure that lifts accuracy by $>76$~pp in the canonical regime and by $>78$~pp in the literature-proxy regime.


% =============================================================================
\section{Implications for the rest of the thesis}
\label{sec:ch1-implications}

This chapter has established the empirical signature and mechanistic interpretation of hardware-instance overfitting.  The remaining chapters build on this foundation in three directions.

\textbf{Chapter~\ref{chap:framework}} constructs the simulation framework that every subsequent experiment inherits: a profile-driven, PyTorch-native behavioural simulator with explicit device-profile substitution and hybrid analogue/digital deployment decomposition.

\textbf{Chapter~\ref{chap:hat-taxonomy}} maps the space of HAT cadences, noise profiles, and transfer guarantees, formalising the block-stationary guarantee that underpins Ensemble HAT.

\textbf{Chapter~\ref{chap:failure-mode-atlas}} catalogues five named, reproducible failure modes that constrain deployment, including the severe-NL regime where the revised gradient-scaling recipe recovers to the $\sim$80--82\% band.

\textbf{Chapter~\ref{chap:mitigation-case-studies}} evaluates the interventions that partially or fully close each gap identified in the atlas.

\textbf{Chapter~\ref{chap:deployment-envelope}} synthesises the evidence into actionable design rules for program managers deciding whether to tape out a ViT accelerator on an organic process.

The central open question is whether the ranking logic and failure-mode boundaries survive the transition to larger vision tasks and to language-model workloads.  Chapter~\ref{chap:outlook} addresses this question through a roadmap of thesis-completion experiments and framework extensions.
