% !TeX program = tectonic
\documentclass[11pt,a4paper]{letter}
\usepackage[margin=1in]{geometry}
\usepackage{amsmath}
\usepackage{enumitem}
\setlist{nosep}
\setlength{\parskip}{0pt}

\signature{Songqiao Li}

\begin{document}

\begin{letter}{Editorial Office\\Nature Electronics\\Springer Nature}

\opening{Dear Editors,}
\vspace{-1em}

\noindent\textbf{Submission:} Hardware-Aware Training and Precision-Retention Frontiers for Low-Precision Analog Vision\\
\textbf{Article Type:} Article

\textbf{Editorial Summary}

We present an open-source simulation framework for organic optoelectronic compute-in-memory inference with modern vision transformers. Our central contribution, Ensemble HAT, resamples device-to-device mismatch masks every epoch. On Tiny-ViT-5M, it raises fresh-instance accuracy from chance level (10.00\%) to 86.16\% (mean $\pm$ 0.19\%), enabling robust deployment without per-device calibration.

Post-fix severe-NL retraining recovers to the $\sim$80--82\% band, falsifying the previously reported $\sim$30\% ceiling and localizing it to a resolved software artifact. A residual gap of roughly 15~pp to the digital baseline remains, whose origin---physical limit vs algorithmic frontier---awaits higher-order surrogates or measured device calibration.

\textbf{Novelty Claims}
\begin{itemize}
    \item We introduce the first open-source framework for organic optoelectronic CIM with vision transformers, providing PyTorch-native device-profile substitution and hybrid analog/digital deployment.
    \item We identify hardware-instance overfitting in organic CIM inference and propose Ensemble HAT, a training-time mismatch-mask resampling fix that restores generalization.
    \item Post-fix severe-NL retraining recovers to the $\sim$80--82\% band, falsifying the previously reported $\sim$30\% ceiling, with a residual gap whose closure awaits higher-order surrogates or measured device data.
\end{itemize}

\textbf{Audience}

Nature Electronics readers will find a reusable methodology for evaluating organic device assumptions on vision tasks and a rigorous negative-result anchor that delineates the limits of first-order behavioral simulation.

\textbf{Related Submissions}

This work is not under consideration elsewhere.

\textbf{Reviewer Information}

Suggested and excluded reviewers are entered through the submission system.

\textbf{Contact Information}

Li Songqiao / 2622507532@qq.com

\closing{Sincerely,}

\end{letter}
\end{document}
