% =============================================================================
% Acknowledgements, Dedication, and Declaration
% =============================================================================

\chapter*{Dedication}
\addcontentsline{toc}{chapter}{Dedication}

\vspace*{2cm}
\begin{center}
\textit{To my family,\\
whose unwavering belief in me has been the quiet foundation upon which\\
every ambition of mine has been built.}
\end{center}

\vspace*{1cm}
\begin{center}
\textit{This work is dedicated to all those who reminded me,\\
in moments of doubt, that the pursuit of knowledge is never walked alone.}
\end{center}

\clearpage

% =============================================================================
\chapter*{Acknowledgements}
\addcontentsline{toc}{chapter}{Acknowledgements}

First and foremost, I owe an immense debt of gratitude to my advisor for intellectual generosity, patience, and steady guidance throughout this work. I am also thankful to the members of my committee for their critiques, questions, and breadth of perspective.

This research would not have been possible without collaborators who contributed expertise at critical junctures. I am grateful for discussions on device physics, code review, software engineering, cross-layer simulation methodology, and the many technical exchanges that clarified the scope of this thesis.

The computational experiments presented in this thesis were enabled by access to high-performance computing resources. I gratefully acknowledge the NVIDIA Apamayo cluster and the computing infrastructure that made large-scale neural architecture search and co-design studies feasible. I also extend my thanks to the system administrators and support staff who maintained these resources with professionalism and responsiveness.

Financial and institutional support will be listed in the official submission metadata. Their investment in fundamental research made this dissertation possible, and I am honored to have been supported by that research environment.

Finally, to my family, friends, and partner: thank you for bearing with the absences, the late nights, and the occasional obsession. Your emotional support was not incidental to this work---it was essential. You reminded me that there is a world beyond the terminal, and that perspective kept me whole.

\clearpage

% =============================================================================
\chapter*{Declaration}
\addcontentsline{toc}{chapter}{Declaration}

I hereby declare that this thesis is my own original work and has not been submitted for any other degree or qualification at this or any other institution. All sources used in the preparation of this thesis have been duly acknowledged. Where the work of others is referenced or built upon, proper attribution has been given. Any errors or omissions remain my own.

\vspace{2cm}
\noindent
Li Songqiao\\
City and date to be confirmed for official submission

% =============================================================================
\endinput
