% !TeX program = tectonic
\documentclass[12pt]{report}

% ---------------------------------------------------------------
% Geometry and layout (thesis format: generous margins)
% ---------------------------------------------------------------
\usepackage[margin=1.25in]{geometry}

% ---------------------------------------------------------------
% Essential packages (adapted from manuscript preamble)
% ---------------------------------------------------------------
\usepackage{graphicx}
\graphicspath{{../latex_gpt/figures/}}
\usepackage{booktabs}
\usepackage{array}
\usepackage{tabularx}
\usepackage{amsmath,amssymb}
\usepackage{algorithm}
\usepackage{algpseudocode}
\usepackage{microtype}
\usepackage{tikz}
\usetikzlibrary{arrows.meta,positioning,fit,calc}
\usepackage{subcaption}
\usepackage[font=small,labelfont=bf]{caption}
\usepackage{placeins}
\usepackage{enumitem}
\usepackage[super,sort&compress]{natbib}
\usepackage[colorlinks=true,linkcolor=blue,citecolor=blue,urlcolor=blue]{hyperref}
\usepackage{cleveref}

% ---------------------------------------------------------------
% Float and list spacing tweaks (from manuscript)
% ---------------------------------------------------------------
\setlength{\textfloatsep}{10pt plus 2pt minus 2pt}
\setlength{\floatsep}{8pt plus 2pt minus 2pt}
\setlength{\intextsep}{8pt plus 2pt minus 2pt}
\setlength{\abovecaptionskip}{5pt}
\setlength{\belowcaptionskip}{0pt}
\setlength{\emergencystretch}{3em}
\setlist[enumerate]{topsep=3pt,itemsep=2pt,leftmargin=1.4em}

% ---------------------------------------------------------------
% Custom macros (from manuscript)
% ---------------------------------------------------------------
\newcommand{\SuppFigNoiseSweep}{S3}
\newcommand{\SuppFigZeroShot}{S4}
\newcommand{\SuppFigRetention}{S6}
\newcommand{\SuppFigFrontend}{S7}
\newcommand{\SuppFigSnr}{S8}
\newcommand{\SuppTableFrontend}{S5}
\newcommand{\SuppTableFlowers}{S7}

% ---------------------------------------------------------------
\newcommand{\ThesisTitle}{Profile-Driven Simulation and Hardware-Aware Training for Organic Optoelectronic Compute-in-Memory Vision}

% Title-page metadata. Formal advisor/department fields still need user confirmation.
% ---------------------------------------------------------------
\title{\textbf{\ThesisTitle}}
\author{Li Songqiao}
\date{May 2026}

% ---------------------------------------------------------------
% Document
% ---------------------------------------------------------------
\begin{document}

% Title page
\begin{titlepage}
    \centering
    \vspace*{2cm}
    {\Huge\bfseries \ThesisTitle\par}
    \vspace{2cm}
    {\LARGE by\par}
    \vspace{1cm}
    {\LARGE Li Songqiao\par}
    \vspace{2cm}
    {\Large Master's thesis draft\par}
    \vspace{2cm}
    {\Large Formal advisor, department, degree, and university metadata\\to be confirmed before official submission.\par}
    \vfill
    {\Large May 2026\par}
\end{titlepage}

\chapter*{Abstract}
\addcontentsline{toc}{chapter}{Abstract}
Compute-in-memory (CIM) architectures reduce the data-movement bottleneck of conventional digital systems by performing matrix--vector multiplications directly inside memory arrays. Among emerging device routes, organic optoelectronic devices offer a distinctive physical substrate for neuromorphic vision systems through multi-level conductance tunability, low-voltage operation, and intrinsic photosensitivity. Translating these device-level advantages into system-level visual performance, however, remains constrained by device variability, limited analog-to-digital converter (ADC) resolution, retention drift, and transfer across fresh hardware instances.

This thesis develops a reusable hardware-aware training (HAT) simulation framework for organic optoelectronic CIM arrays supporting vision transformers and convolutional neural networks under realistic device non-idealities. The framework maps literature-reported device characteristics, including conductance window, cycle-to-cycle variability, device-to-device mismatch, retention dynamics, photoresponse nonlinearity, and write nonlinearity, into task-level visual accuracy.

The experiments first evaluate ResNet-18, ConvNeXt-Tiny, and Tiny-ViT-5M across CIFAR-10, CIFAR-100, Flowers-102, and ImageNet-1k baselines. Building on these baselines, the thesis introduces a taxonomy of HAT regimes and identifies hardware-instance overfitting: standard fixed-mask HAT can perform well on the training-time hardware instance yet collapse to near-chance accuracy on fresh mismatch realizations. Ensemble HAT addresses this failure by resampling structured device-to-device mismatch fields at epoch boundaries, recovering CIFAR-10 fresh-instance accuracy to \(86.16\pm0.19\%\) across three training seeds.

The thesis further constructs a failure-mode atlas for organic optoelectronic CIM deployment, covering hardware-instance overfitting, severe write-nonlinearity collapse, the limited diagnostic ceiling of MLP-only linearization, bounded degradation under spatially correlated mismatch, and retention-drift plateaus. These results provide system-level decision support for algorithm--device co-design and clarify which training interventions transfer to new hardware instances and which remain constrained by physical modeling assumptions.

% Table of contents
\tableofcontents

% ---------------------------------------------------------------
% Chapters
% ---------------------------------------------------------------
\include{chapter_1_hat_instance_overfitting}
\include{chapter_2_framework}
\include{chapter_3_hat_taxonomy}
\include{chapter_4_failure_modes}
\include{chapter_5_mitigation}
\include{chapter_6_physical_realism}
\include{chapter_7_deployment}
\include{chapter_8_outlook}

% ---------------------------------------------------------------
% Bibliography
% ---------------------------------------------------------------
\bibliographystyle{unsrtnat}
\bibliography{../latex_gpt/refs_gpt}

% ---------------------------------------------------------------
% Appendix (optional)
% ---------------------------------------------------------------
% \appendix
% \include{appendix_a}
% Uncomment the lines above when appendix files are ready.

\end{document}
