% !TeX program = tectonic
\documentclass[11pt,a4paper]{letter}
\usepackage[margin=1in]{geometry}
\usepackage{amsmath}
\usepackage[colorlinks=true,linkcolor=blue,citecolor=blue,urlcolor=blue]{hyperref}
\usepackage{enumitem}
\setlist{nosep}

\signature{Songqiao Li}

\begin{document}

\begin{letter}{Editorial Office\\Nature Electronics\\Springer Nature}

\opening{Dear Editors,}

\textbf{Submission:} Hardware-Instance Overfitting in Organic Optoelectronic Compute-in-Memory: A Profile-Driven Simulation and Mitigation Framework

\textbf{Article Type:} Article

\textbf{Reviewer Information:} Suggested and excluded reviewers will be provided through the submission system.

\vspace{0.5em}
\textbf{Editorial Summary}

We present a first-order behavioural simulation framework that bridges organic-device literature metrics to task-level vision accuracy, exposing a previously unnamed failure mode---hardware-instance overfitting---and providing a principled treatment. Our central result is Ensemble Hardware-Aware Training (HAT), a distribution-matching objective that resamples device-to-device (D2D) mismatch masks each training epoch. Ensemble HAT raises fresh-instance accuracy on unseen hardware arrays from chance level (10.00\%) to 86.37$\pm$1.54\% on Tiny-ViT-5M (CIFAR-10), a gain of more than 76 percentage points. The protocol is theoretically grounded as an implicit gradient-$L_2$ regulariser equivalent to dropout-as-$L_2$ (Wager 2013), derived in Supplementary Note~S-Theory.

The framework is validated across three scenario pillars: (i) canonical uniform noise (10\% D2D, 5\% C2C, $NL=1.0$), where Ensemble HAT recovers 86.37$\pm$1.54\%; (ii) a literature-anchored OPECT organic-phototransistor proxy ($\sigma_{\mathrm{D2D}}=3\%$, $\sigma_{\mathrm{C2C}}=2\%$), where zero-shot transfer reaches 88.53$\pm$0.08\% without retraining; and (iii) severe nonlinear write ($NL=2.0$), where hardware-aware training with a revised gradient-scaling recipe recovers to the $\sim$80--82\% band. A residual gap of roughly 15~pp to the 98.06\% digital baseline remains, whose closure awaits higher-order gradient corrections or device-level linearisation.

To bound realistic deployment performance, we report a dual-column ADC discipline: ADC-off training-surrogate baselines are paired with post-module-output hook diagnostic 8-bit ADC-on evaluations calibrated per-instance on each fresh hardware array. Under severe NL, the ADC-on headline differs from the ADC-off surrogate by only $-0.10$~pp on average, while a 6-bit spot-check reveals a larger $-2.8$~pp cliff. Stage-2 per-instance calibration confirms the ADC-on headline is stable to within $0.01$~pp run-to-run std, validating the static-calibration baseline. Spatially correlated D2D is also evaluated under a separable AR(1) proxy: moderate correlation ($\rho=0.3$) degrades Ensemble HAT by only $-1.76$~pp (84.57$\pm$2.39\%), and stronger correlation ($\rho=0.5$) by $-4.20$~pp (82.12$\pm$3.95\%), with no instance collapsing below 73.7\%---empirical evidence of bounded degradation rather than structural collapse (canonical evaluation protocol, $NL=1.0$, AMP disabled).

\vspace{0.5em}
\textbf{Why Nature Electronics?}

This work sits at the intersection of emerging device physics, mixed-signal system design, and machine-learning robustness. Organic optoelectronic compute-in-memory offers multi-modal tuning and excellent low-power analogue behaviour, yet lacks a systematic methodology for translating literature device parameters into task-level accuracy predictions before fabrication closure. We provide that methodology: a PyTorch-native, profile-driven behavioural simulator with explicit device-profile substitution and hybrid analogue/digital deployment decomposition. The Ensemble HAT contribution is not merely a training trick; it is a distribution-matching objective with theoretical grounding that generalises across noise profiles, spatial correlation structures, and front-end nonlinearities. We believe this hardware--algorithm co-design narrative, combined with strict source-data provenance and independent cross-checks, is a natural fit for Nature Electronics' readership.

\vspace{0.5em}
\textbf{Methodological Rigor}

All numerical claims in the manuscript carry source-data provenance distinguishing canonical evaluations, diagnostic ablations, and supplementary sensitivity checks. The severe-NL narrative uses a revised gradient-scaling recipe and reports recovery-band evidence under matched training and evaluation settings. The framework is further fabrication-ready: measured device statistics from a fabricated organic optoelectronic array can be integrated via the ingest pipeline described in Supplementary Note~S-HW as soon as such measurements become available.

\vspace{0.5em}
\textbf{Companion Work and Collaboration}

A companion paper will extend the framework to the attention-pathway analog-CIM regime, specifically KV-cache storage in organic crossbar arrays; the outlook section previews this extension and demonstrates that the diagnostic-treatment signature generalises beyond the feedforward path. We are also collaborating with a fabrication team to obtain measured D2D spatial covariance kernels and retention trajectories; these will be integrated in revision stage via the profile-fitting pipeline.

\vspace{0.5em}
\textbf{Key Contributions}

\begin{enumerate}
    \item Profile-driven, PyTorch-native behavioural simulator for organic optoelectronic CIM with explicit device-profile substitution and hybrid analogue/digital deployment decomposition
    \item Ensemble HAT: theoretically grounded distribution-matching training raises fresh-instance accuracy from 10.00\% to 86.37$\pm$1.54\%
    \item Literature-anchored OPECT case study (88.53$\pm$0.08\% zero-shot) and severe-NL recovery to the $\sim$80--82\% band under revised training
    \item Dual-column ADC discipline (surrogate + hook-diagnostic) that bounds realistic deployment performance before silicon validation
    \item Correlated-D2D bounded-degradation evidence under AR(1) spatial structure, with worst-case instance 73.7\% at $\rho=0.5$
\end{enumerate}

\vspace{0.5em}
\textbf{Transparency Disclosures}

\begin{itemize}
    \item \textbf{Preprint:} None; submitted exclusively to Nature Electronics.
    \item \textbf{Prior Publication:} None; the framework was developed for this submission.
    \item \textbf{Code Availability:} A reviewer-accessible archive will be provided at submission; a public release will follow at acceptance.
    \item \textbf{Competing Interests:} The authors declare no competing interests.
\end{itemize}

The submission package includes a main manuscript and a supplementary information PDF. All datasets are publicly available, and all key claims carry error bars from multiple Monte Carlo runs or training seeds.

\closing{Sincerely,}

\end{letter}
\end{document}
